Kindly Bent to Free Us

### 2018-11-09 kindly/infer/infer.tex
Add normalization procedure.

We now formulate type inference for the \lang language. Our inference technique is based on the HM(X) framework~\citep{DBLP:journals/tapos/OderskySW99} which
presents how to merge Hindley-Milner type inference for let-polymorphism and
constraint solving. We first present our constraint language, we then show
our slight extension of HM(X) that support a kind system and affine types.

@@

We consider $\mathcal C$ the set of constrains.
The grammar of constrains, presented
in \cref{grammar:constraint}, follows the traditional HM(X) formulation
with conjunctions,projections and type inequalities. The only new
element specific to our approach are kind equalities.
Entailment is noted $\entail{C}{D}$, where $D$ is a consequence of the
constraints $C$. The entailment rules are provided in \cref{rules:entail}.

@@

\begin{figure*}[h]
  \input{infer/entails}
  \caption{Entailment rules -- $\entail{C}{D}$ }
  \label{rules:entail}
\end{figure*}

\subsubsection{Solved forms}

The set of solved forms, noted $\mathcal S$,
is the set of constraints than can be used inside
type and kind schemes. $\mathcal S$ is composed only of kind
inequalities \emph{over variables}. For convenience, if $C\in\mathcal S$, we
note $C$ as a list of kind inequalities: $\Cleq{\kvar_i}{\kvar_{i'}}^n$.
\TODO{Extend the properties of solved forms}

We consider the existence of a function $\normalize$ which takes
a constraint in $\mathcal C$ and a unifier $\psi$ and returns a constraint
in solved form $C'$, and an updated unifier.

@@

\TODO{Explain $\Sigma$}

@@

\TODO{Explain more?}

@@

\subsection{Constraint normalization and generalization}

>>>>

We now formulate type inference for the \lang language. Our inference technique is based on the HM(X) framework~\citep{DBLP:journals/tapos/OderskySW99} which
presents how to implement Hindley-Milner type inference in a language
with qualified types. We first present various preliminaries definitions,
we then show our extension of HM(X) that support a kind system and affine
types.

@@

We consider $\mathcal C$ the set of constrains.
The grammar of constrains, presented
in \cref{grammar:constraint}, follows the traditional HM(X) formulation
with conjunctions,projections and type inequalities. The only new
element specific to our approach are kind equalities.
Entailment is noted $\entail{C}{D}$, where $D$ is a consequence of the
constraints $C$.
The base entailment rules are shown in \cref{rules:entail}, we consider
the completion of that system to form a cylindric constraint system.
\TODO{Give the properties}

We say that $C$ and $D$ are equivalent, noted $C \equivC D$,
when $\entail{C}{D}$ and $\entail{D}{C}$.

@@

\begin{figure*}[h]
  \input{infer/entails}
  \caption{Base entailment rules -- $\entail{C}{D}$ }
  \label{rules:entail}
\end{figure*}

We note $\mathcal S$ the set of solved forms
which can be used inside type and kind schemes.
We define $\mathcal S$ as $\mathcal C$ quotiented by the relation $\equivC$.
%
We consider the existence of a function $\normalize$ which takes
a constraint in $\mathcal C$ and a unifier $\psi$ and returns a constraint
in solved form $C' \in \mathcal S$,
and an updated unifier. We detail the implementation
of the normalization function in \cref{sec:normalize}

% $\mathcal S$ is composed only of kind
% inequalities \emph{over variables}. For convenience, if $C\in\mathcal S$, we
% note $C$ as a list of kind inequalities: $\Cleq{\kvar_i}{\kvar_{i'}}^n$.
% \TODO{Extend the properties of solved forms}

@@

In order to determine if a variable is used in an affine manner, we must track
it's uses and the associated kinds. For instance, in the expression
$(x,x)$, $x$ is used twice. If $x$ is of type $\tau$, which is of kind $k$,
we must add the constraint $\Cleq{k}{\kun}$. In order to record uses and kinds,
our type inference relation will return a ``usage map'', noted $\Sv$,
which associates each variable to a list of kinds corresponding to each use site
of the variable.
We note $\Sone{x}{k_1,\dots,k_n}$ the singleton usage map.
Common operations on usages maps are defined in \cref{op:usgmap}. We define
regular set operations such as union
($\Sv\cup\Sv'$) and intersection ($\Sv\cap\Sv'$), along
with constraint creation $\Cleq{\Sv}{k_0}$ which indicates that all the kinds
presents in the usage map $\Sv$ are constrained by the kind $k_0$.
These operations corresponds roughly to the various parts of the join
operator on sets $\esplit{}{}$.

\begin{figure}[h]\centering
\begin{align*}
  \Sv\cap\Sv'
  &= \left\{ x \to l,l'
    \mid \Sone{x}{l}\in\Sv \text{ and } \Sone{x}{l'}\in\Sv' \right\} \\
  \Sv\cup\Sv'
  &= \Sv\cap\Sv' \sqcup
    (\Sv\Sdel{x' \mid x'\in\Sv'}) \sqcup
    (\Sv'\Sdel{x \mid x\in\Sv})
  &\text{where $\sqcup$ is the disjoint union}\\
  \Cleq{\Sv}{k_0}
  &= \bigwedge_{\Sone{x}{l_x}\in\Sv} \bigwedge_{k\in l_x} \Cleq{k}{k_0}
\end{align*}
\caption{Operation on usage maps $\Sv$}
\label{op:usgmap}
\end{figure}

@@

@@

\subsection{Normal forms and generalization}
\label{sec:normalize}

\newcommand\lub\bigvee
\newcommand\glb\bigwedge

We now place ourselves in a more general context where kinds are either variables
or constants belonging to a total bounded lattice $(\mathcal L, \leq)$ ie.,
a lattice which admits a total order and upper and lower bounds.
We note lattice elements $l$ and $\glb_i l_i$ (resp. $\lub_i l_i$)
the greatest lower bound (resp. least upper bound) in $\mathcal L$.

The normalization procedure is noted $\normalize{C_0}{\unif_0}$ where
$C_0\in \mathcal C$ is a set of constraints and $\unif_0$ is a unifier.
It returns a constraint $C \in \mathcal S$ in
solved form and a unifier $\unif$.
Our unification algorithm proceed as follows:
\begin{enumerate}
\item Solve all type equality constraints through Herband unification and
  gather all existential quantifications at the front of the constraint.
  We obtain a constraint $C = \exists \kvar_i,\ \Cleq{k_j}{k'_j}_j$ and
  a unifier $\unif$.

  We note $\mathcal R$ the relation $\Cleq{k_j}{k'_j}_j$,
  $\mathcal G$ the underlying directed graph and $V$ its vertices.

\item Saturate the lattice equalities in $\mathcal R$.

  More precisely, for each kind variable $\kvar \in V$,
  for each constant $l_i$ (resp. $l_j$) such that
  there is a path from $l_i$ to $\kvar$ (resp. from $\kvar$ to $l_j$) in $\mathcal G$,
  add an edge from $\lub l_i$ to $\kvar$
  (resp. from $\kvar$ to $\glb l_j$).
  This step is well defined since
  our lattice is bounded and $\lub\emptyset$ and $\glb\emptyset$ are well defined.

  We also complement $\mathcal R$ with $(\leq)$ by adding an edge
  between related constants.
\item For each strongly connected component in $\mathcal G$, unify all its vertices and replace it by a representative and complete the unifier accordingly.
  The representative of a strongly connected component $g$ can be determined as follows:
  \begin{itemize}
  \item If $g$ does not contain any constant, then the representative
    is a fresh kind variable.
  \item If $g$ contains exactly one constant, it is the representative.
  \item Otherwise, the initial constraint $C_0$ is not satisfiable.
  \end{itemize}
  Note that this step will also detect all unsatisfiable constraints.
\item Take the transitive closure of $\mathcal R$.
\item Remove all the vertices corresponding to the kind variables $\kvar_i$
  that are existentially quantified in $C$.
\item Remove the extremums of $\mathcal L$ and the edges of $(\leq)$
  from $\mathcal R$.
\item Take the transitive reduction of $\mathcal R$.
  % Return $C = \left\{ k \leq k' \mid k \operatorname{\mathcal R}k' \right\}$
\end{enumerate}

\TODO{Add an example}

\begin{lemma}
  Given a constraint $C\in\mathcal C$, a substitution $\unif$ and
  $(C',\unif') = \normalize{C}{\unif}$, we have the following properties:
  \begin{enumerate}
  \item $C'$ is of the form $\{ k_i \leq k'_i\}_i$.
  \item For any disjoint subformulas $C_1$ and $C_2$ of $C'$,
    then $C_1 \operatorname{\nvdash_e} C_2$ and $C_2 \operatorname{\nvdash_e} C_1$
  \item For each $\kvar \in \operatorname{fv}(C)$,
    there exists two constants $l_\bot,l_\top \in \mathcal L$  such that
    $\entail{C}{\Cleq{l_\bot}{\kvar}}$ and $\entail{C}{\Cleq{\kvar}{l_\top}}$.
  \end{enumerate}
\end{lemma}
% \begin{proof}

%   \begin{enumerate}
%   \item Directly follows from step (1) and (5).
%   \end{enumerate}
%   (1) is a direct consequence of step 
% \end{proof}

\begin{lemma}[Normal form]
  Given a constraint $D\in\mathcal C$, a substitution $\phi$ and
  $(C,\unif) = \normalize{D}{\phi}$,
  then $\phi\leq\unif$,
  $C \equivC \unif D$ and
  $\unif C = C$.
\end{lemma}
\begin{proof}
  \TODO{}
\end{proof}

\begin{lemma}[Uniqueness]
  Given $(C_1,\unif_1)$ and $(C_2,\unif_2)$ such that
  $\unif_1 C_1 \equivC \unif_2 C_2$, then
  $\normalize{C_1}{\unif_1}$ and $\normalize{C_2}{\unif_2}$
  are identical up to $\alpha$-renaming.
\end{lemma}
\begin{proof}
  \TODO{}
\end{proof}

\begin{theorem}
  The procedure $\operatorname{normalize}$ is complete and computes
  unique principal normal forms for our constraint system.
\end{theorem}
